smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-10-16 manuscript/manuscript.qmd
Implemented SciComms and FDA review feedback, final version for initial submission

[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>. The R package presented in this study can be downloaded from CRAN via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.

@@

Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their capture of clinical details, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

@@

-   **Missing not at random (MNAR)**: The missingness depends on unoberserved covariate(s). For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~). The missingness can also just depend on the actual value of the partially observed covariate itself (MNAR~value~).
:::

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure quality, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website: <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is (globally) MCAR.

>>>>

[**Analytical code and data sharing statement:**]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All R code, materials and depedencies can be found at <https://gitlab-scm.partners.org/drugepi/smdi-manuscript>. The R package presented in this study can be downloaded from the comprehensive R archive network (CRAN) via `install.packages("smdi")` or from [https://janickweberpals.gitlab-pages.partners.org/smdi](https://janickweberpals.gitlab-pages.partners.org/smdi/){.uri}.

@@

Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their capture of clinical details, EHR can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

@@

-   **Missing not at random (MNAR)**: The missingness depends on unobserved covariate(s). For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~). The missingness can also just depend on the actual value of the partially observed covariate itself (MNAR~value~).
:::

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on the comprehensive R archive network (<https://cran.r-project.org/web/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via `install.packages("smdi")`. To ensure quality, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment. Additional resources such as documentation and vignettes are provided on the package website: <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test, which means that `smdi_hotelling()` returns a test statistic and p-value for each POC. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data are (globally) MCAR.
